---
title: "Can we predict cases of dengue with climate variables?"
author: "Francisco Requena"
date: '2017-12-09'
slug: can-we-predict-cases-of-dengue-with-climate-variables
tags: []
categories:
- R
- machine learning
---


Recently, I discovered a new website about competitions that it is not called Kaggle! Its name is Drivendata.

**DrivenData** offers different competitions related with multiple types of field, such as health (oh yes!), ecology, society… with a common element: to face the world’s biggest social challenges.

I decided to join my first competition called *‘DengAI: Predicting Disease Spread‘*. In this case, the user receives a set of weather information (temperatures, precipitations, vegetations) from two cities: **San Juan** (Puerto Rico) and **Iquitos** (Peru) with total cases of dengue by year and week of every year.

The goal of the competition is to develop a prediction model that would be able to anticipate the cases of dengue in every city depending on a set of climate variables.

The DrivenData’s blog wrote some days ago, a post about a fast approach with this dataset. It was written in Python. So, I decided to “translate” to R language.

The next code is divided into three main points:


**1.** Code with clean tasks (transform NA values, remove of columns…) and exploratory analyses.

**2.** Function with every step during cleaning of data.

**3.** Development of model, prediction and comparison of predicted vs real total cases detected.




```{r, normal, message= FALSE}
# Load of libraries

library(tidyverse)
library(zoo)
library(corrplot)
library(MASS)
library(reshape2)
```

```{r}
# Load data
train_features <- read.csv('data/dengue_post/dengue_features_train.csv')
  
train_labels <- read.csv('data/dengue_post/dengue_labels_train.csv')

test_features <- read.csv('data/dengue_post/dengue_features_test.csv')

submission_format <- read.csv('data/dengue_post/submission_format.csv')
  
# Filter of data by city  

sj_train_labels <- filter(train_labels, city == 'sj')
sj_train_features <- filter(train_features, city == 'sj')

iq_train_labels <- filter(train_labels, city == 'iq')
iq_train_features <- filter(train_features, city == 'iq')
```


```{r}
# Is there NA values?

df_na_sj <- as.data.frame(apply(sj_train_features,2, function(x) any(is.na(x))))
colnames(df_na_sj) <- 'is_there_NA'
df_na_sj$number_NA <- apply(sj_train_features,2, function(x) sum(is.na(x)))
df_na_sj$mean_NA <- apply(sj_train_features, 2, function(x) mean(is.na(x)))

df_na_iq <- as.data.frame(apply(iq_train_features, 2, function(x) any(is.na(x))))
colnames(df_na_iq) <- 'is_there_NA'
df_na_iq$number_NA <- apply(iq_train_features, 2, function(x) sum(is.na(x)))
df_na_iq$mean_NA <- apply(iq_train_features, 2, function(x) mean(is.na(x)))
```



```{r}

# Vegetation Index over Time Plot with NAs

ggplot(sj_train_features, aes(x = as.Date(week_start_date), y = ndvi_ne )) +
  ggtitle('Vegetation Index over Time') +
  theme_bw() +
  xlab('Title') +
  geom_line(na.rm = FALSE, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r, message=FALSE, warning=FALSE}

# Remove 'weekofyear' column

sj_train_features <- dplyr::select(sj_train_features, -week_start_date)

iq_train_features <- dplyr::select(iq_train_features, -week_start_date)

# Fill the NA values with the previous value

sj_train_features <- sj_train_features %>%
            do(na.locf(.))

iq_train_features <- iq_train_features %>%
            do(na.locf(.))

# Distribution of labels

# print(mean(sj_train_labels$total_cases))
# print(var(sj_train_labels$total_cases))
# 
# print(mean(iq_train_labels$total_cases))
# print(var(iq_train_labels$total_cases))


ggplot(sj_train_labels, aes(x = total_cases)) +
  theme_bw() +
  ggtitle('Cases of dengue in San Juan') +
  geom_histogram() +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(iq_train_labels, aes(x = total_cases)) +
  theme_bw() +
  ggtitle('Cases of dengue in Iquitos') +
  geom_histogram() +
  theme(plot.title = element_text(hjust = 0.5))

# Add total_cases column to *_train_features dataframes


# sj_train_features <- left_join(sj_train_features, sj_train_labels, by = c('city', 'year', 'weekofyear'))
 sj_train_features$total_cases <- sj_train_labels$total_cases

# iq_train_features <- left_join(iq_train_features, iq_train_labels, by = c('city', 'year', 'weekofyear'))
iq_train_features$total_cases <- iq_train_labels$total_cases

# Correlation matrix

m_sj_train_features <- data.matrix(sj_train_features)
m_sj_train_features <- cor(x = m_sj_train_features[,3:24], use = 'complete.obs', method = 'pearson')

m_iq_train_features <- data.matrix(iq_train_features)
m_iq_train_features <- cor(x = m_iq_train_features[,3:24], use = 'everything', method = 'pearson')

# Correlation Heatmap

corrplot(m_sj_train_features, type = 'full', tl.col = 'black', method="shade")

corrplot(m_iq_train_features, type = 'full', tl.col = 'black', method = 'shade')

# Correlation Bar plot

df_sj_train_features <- data.frame(m_sj_train_features)[2:21,] 
df_sj_train_features <- dplyr::select(df_sj_train_features, total_cases) 
                                    
df_iq_train_features <- data.frame(m_iq_train_features)[2:21,]
df_iq_train_features <- dplyr::select(df_iq_train_features, total_cases) 

ggplot(df_sj_train_features, aes(x= reorder(rownames(df_sj_train_features), -total_cases), y = total_cases)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  ggtitle('Correlation of variables in San Juan') +
  ylab('Correlation') +
  xlab('Variables') +
  coord_flip()

ggplot(df_iq_train_features, aes(x= reorder(rownames(df_sj_train_features), -total_cases), y = total_cases)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  ggtitle('Correlation of variables in Iquitos') +
  ylab('Correlation') +
  xlab('Variables') +
  coord_flip()
```

```{r, message=FALSE, warning=FALSE}
# Function data cleaning

data_clean <- function(df_dengue_features, df_dengue_labels = NULL, add_cases = TRUE) {
  
  # Filter by city
  sj_df_dengue_features <- filter(df_dengue_features, city == 'sj')
  iq_df_dengue_features <- filter(df_dengue_features, city == 'iq')
  
  if (add_cases == TRUE) {
  sj_df_dengue_labels <- filter(df_dengue_labels, city == 'sj')
  iq_df_dengue_labels <- filter(df_dengue_labels, city == 'iq')
  }
  # Removing week_start_date column
  sj_df_dengue_features <- dplyr::select(sj_df_dengue_features, -week_start_date)
  iq_df_dengue_features <- dplyr::select(iq_df_dengue_features, -week_start_date)

  # Fill of NA values with the previous value
  sj_df_dengue_features <- sj_df_dengue_features %>%
    do(na.locf(.))
  
  iq_df_dengue_features <- iq_df_dengue_features %>%
    do(na.locf(.))
  
  # Add total_cases to dataframe with features
  if (add_cases == TRUE) {
  sj_df_dengue_features$total_cases <- sj_df_dengue_labels$total_cases
  iq_df_dengue_features$total_cases <- iq_df_dengue_labels$total_cases
  }
  
  # Converting character columns into numbers
  sj_df_dengue_features <- as.data.frame(apply(sj_df_dengue_features,2,as.numeric))
  sj_df_dengue_features$city <- rep('sj', nrow(sj_df_dengue_features))
  iq_df_dengue_features <- as.data.frame(apply(iq_df_dengue_features,2,as.numeric))
  iq_df_dengue_features$city <- rep('iq', nrow(iq_df_dengue_features))
  
  result <- list(sj_df_dengue_features, iq_df_dengue_features )
  
  return(result)
}
```

```{r, message=FALSE, warning=FALSE}
# Getting data_training clean

data_train <- data_clean(train_features, train_labels, TRUE)

# Getting negative binomials models by city

training_sj <- glm.nb(formula = total_cases ~ reanalysis_specific_humidity_g_per_kg +
                     reanalysis_dew_point_temp_k +
                     station_min_temp_c +
                     station_avg_temp_c, data = data_train[[1]])

training_iq <- glm.nb(formula = total_cases ~ reanalysis_specific_humidity_g_per_kg +
                        reanalysis_dew_point_temp_k +
                        station_min_temp_c +
                        station_avg_temp_c, data = data_train[[2]])

# Getting data_test clean

data_test <- data_clean(test_features, add_cases = FALSE)


# Testing model with training data

prediction_train_sj <-  predict(training_sj, data_train[[1]], type = 'response')
prediction_train_iq <-  predict(training_iq, data_train[[2]], type = 'response')

df_prediction_train_sj <- data.frame('prediction' = prediction_train_sj, 'actual' = data_train[[1]]$total_cases,
                                     'time' = as.Date(train_features$week_start_date[1:936]))
df_prediction_train_sj <- melt(df_prediction_train_sj, id.vars = 'time')
ggplot(df_prediction_train_sj, aes(x = time, y = value, color = variable)) +
  geom_line() +
  ggtitle('Dengue predicted Cases vs. Actual Cases (City-San Juan) ')

df_prediction_train_iq <- data.frame('prediction' = prediction_train_iq, 'actual' = data_train[[2]]$total_cases,
                                     'time' = as.Date(train_features$week_start_date[937:1456]))
df_prediction_train_iq <- melt(df_prediction_train_iq, id.vars = 'time')
ggplot(df_prediction_train_iq, aes(x = time, y = value, color = variable)) +
  geom_line() +
  ggtitle('Dengue predicted Cases vs. Actual Cases (City-Iquitos) ')

# Prediction of total_cases in the data set

prediction_sj <-  predict(training_sj, data_test[[1]], type = 'response')
prediction_iq <-  predict(training_iq, data_test[[2]], type = 'response')
 
data_prediction_sj <- data.frame('city' = rep('sj', length(prediction_sj) ), 
                                 'total_cases' = prediction_sj, 
                                 'weekofyear' = data_test[[1]]$weekofyear,
                                 'year' = data_test[[1]]$year )

data_prediction_iq <- data.frame('city' = rep('iq', length(prediction_iq) ), 
                                 'total_cases' = prediction_iq,
                                 'weekofyear' = data_test[[2]]$weekofyear,
                                 'year' = data_test[[2]]$year)


  
submission_format$total_cases <- as.numeric(c(data_prediction_sj$total_cases, 
                                                   data_prediction_iq$total_cases))

submission_format$total_cases <- round(submission_format$total_cases, 0)
  
write.csv(submission_format,
          file = 'submission_format_submit.csv', row.names = F)
```

